---
title: "Tidying Data"
author: "Greg Dean"
date: "`r Sys.Date()`"
output: github_document
---
This document demostrates some techniques to organize data as *tidy data*. 

Once your data is organized into a tidy form, you are able to apply the various tools of the tidyverse to find insights in your data.

The examples in this document are adapted from Chapter 9 in the first edition of _R for Data Science_ by Hadley Wickham & Garrett Grolemund.

## Prerequisites
We will be focusing on the *tidyr* package in this example. *Tidyr* provides a variety of tools that help "tidy up" messy (or non-tidy) data sets.

Since *tidyr* is part of the "core tidyverse" we can load the tidyverse package to gain the functionality we need:

```{r, echo = FALSE}
library(tidyverse)
```
## Representation of the Same Data in Different Ways
The following example represents the same underlying data in four different ways.

Each dataset shows the same values of four variables organized in a different way:

1. Country
2. Year
3. Population
4. Cases

```{r, echo = FALSE}
table1

table2

table3

table4a

table4b
```
While these tables are all representations of the same underlying data, they are not equally easy to use.

One dataset, the _Tidy Data Set_ will be much easier to work with inside the tidyverse.

## Three Rules Which Make a Data Set Tidy
[INSERT THE PICTURE FROM PG. 149 HERE]

1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

In the example above, only `table1` is tidy, as it is the only representation where each column is a variable.

Why ensure the data is tidy?

Two main advantages:

1. **Consistency**. If you have a consistent data structure, it's easier to learn the tools that work with it because they have an underlying uniformity.  

2. **Vectors**. Placing variables in columns allows R's vectorized nature to shine. Most built-in functions in the R Programming language work with vectors of values, and tidy data works well with these types of functions.

```{r echo = FALSE}
# Compute rate per 10,000
table1 %>%
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>%
  count(year, wt = cases)

# Visualize changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country))
```
